Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-07-23 ATVA 2025/Final_version/experiments.tex
update the final version
update draft of final version
ATVA25

\begin{figure}[b!]
	\centering
	\vspace*{-0.3cm}
	\includegraphics[height=7cm]{CNN-B-ADV_layer7_comparison_image85.png}.
	\vspace*{-0.4cm}
	\caption{Average uncertainty of pMILP for neurons of the fourth layer of CNN-B-Adv, 
	when selecting between 0 and 30 ReLUs with different scoring functions.}
	\label{fig_table3}
\end{figure}

@@

Overall, {\sf SAS} is more accurate than GS, more accurate than Huang, more accurate than random choice of variables. {\sf SAS} significantly outperforms other solutions, 
with 2 times less binary variables for the same accuracy (10 vs 22) vs {\sf GS} and 6 times vs Huang \cite{DivideAndSlide} (4 vs 24).
Each node of the 4th layer displays a similar pattern, and the pattern is similar for different images (see appendix).

@@

\begin{table}[t!]	
	\centering
	\begin{tabular}{|c||c c|c c|c c|c c|}
		\hline
		  Image 85& \multicolumn{2}{c|}{45 open ReLUs} & \multicolumn{2}{c|}{50 open ReLUs} 
		  & \multicolumn{2}{c|}{55 open ReLUs} & \multicolumn{2}{c|}{60 open ReLUs}\\ 
		 order & distance & time & distance & time & distance & time & distance & time \\
		\hline \hline
		Gurobi & 0.384 & 252s & 0.368&265s  &0.342 & 287s& 0.315  & 306s  \\ 
		{\sf SAS} (static) & 0.384 & 448s & 0.368&759s  &0.365 &971s & 0.355  & 971s \\
		{\sf GS} (static) & 0.384 & {\bf 251s} &0.368 &{\bf 253s}  & 0.342 &{\bf 270s} & 0.315 & {\bf 281s} \\ \hline
	\end{tabular}
	\caption{Comparison of different orderings with selection of ReLUs by {\sf SAS}.
	\vspace{-0.4cm}}
	\label{table.order}
\end{table}

@@

\begin{figure}[t!]
	\hspace*{-0.8cm}
	\includegraphics[scale=0.5]{CNN-B-ADV_layer7_comparison_image85_4methods.png}
	\caption{Distance to verify vs runtime: comparison between {\sf SAS}, GS, full MILP and Huang's method for different number of opened ReLUs / time-outs.}
	\label{fig555}
\end{figure}

{\sf SAS} (even {\sf GS}$=s_{FSB}$) is much more accurate than full MILP for every reasonable time-outs considered, with $>10$ times fastest runtime at better accuracy. The reason is that even the advanced MILP solver Gurobi struggles to sort out all ReLUs. It is thus particulary important to have accurate scoring functions as {\sf SAS}, in order to optimize both accuracy and runtime. Huang \cite{DivideAndSlide} is limited in ReLUs in the previous layer, reason why it is stuck at relatively poor accuracy. The accuracy/runtime curve from {\sf GS} is closer to {\sf SAS} than the number of nodes / accuracy curve of Fig. \ref{fig_table3}. This is because many ReLU nodes deemed important by {\sf GS} are not relevant for accuracy, but they are also not penalizing runtime. Still, {\sf SAS} is faster than {\sf GS} at  every accuracy. Importantly, {\sf SAS} is more deterministic in its runtime given a number of ReLU nodes, with half the variance in runtime over different output neurons compared with {\sf GS} for similar accuracy (see table \ref{SAS_GS_similarTO} in Appendix), which helps setting a number of ReLU nodes to open.

@@

@@

\begin{table}[t!]
	\centering
	\begin{tabular}{||l||c|c|c||c||c||}
		\hline \hline
		 & $\alpha,\beta$-CROWN & $\alpha,\beta$-CROWN & $\alpha,\beta$-CROWN & Refined & \bf Hybrid \\ 
		 Network & TO=10s & TO=30s & TO=2000s & $\beta$-CROWN & \bf MILP \\ 
		\hline
		MNIST 5$\times$100 & 57\% (6.9s) & 55\% (18.9s) & 50\% (1026s) & 13\% 
		(92s) & 13\% \bf (46s) \\ \hline
		MNIST 5$\times$200 & 50\% (6.5s) & 47\% (17s) & 46\% (930s) & 9\% (80s) & \bf 8\% (71s) \\ \hline
		MNIST 8$\times$100 & 63\% (7.2s) & 58\% (20s) & 58\% (1163s) & 21\% (102s) & \bf 15\% (61s) \\ \hline
		MNIST 8$\times$200 & 56\% (6.8s) & 55\% (18s) & 54\% (1083s) & 16\% (83s) & \bf 8\% (78s) \\ \hline
		MNIST 6$\times$500 & 53\% (6.4s) & 51\% (16s) & 50\% (1002s) & $-$ & 
		\bf 10\% (402s) \\ \hline
		CIFAR CNN-B-Adv & 28\% (4.3s) & 22\% (8.7s) & 20\% (373s) & $-$ & \bf 11\% (417s) \\ \hline \hline
		CIFAR ResNet & 0\% (2s) & 0\% (2s) & 0\% (2s) & $-$ & 0\% (2s) \\ \hline \hline
	\end{tabular}
	\caption{Undecided images ($\%$, {\em lower is better}) as computed by $\alpha,\beta$-CROWN, Refined $\beta$-CROWN and Hybrid MILP on 7 DNNs (average runtime per image). The 6 first DNNs are hard instances. The last DNN (ResNet) is an easy instance (trained using Wong to be easy to verify), provided for reference.
	\vspace{-1cm}}
	\label{table_hybrid}
	\end{table}

@@

\begin{table}[t!]
	\centering
	\begin{tabular}{||c||c|c|c|c|c||}
		\hline
		Upper Bound & SDP-FO & PRIMA & refined $\beta$-CROWN & $\beta$-CROWN & GCP-CROWN \\  \hline
		65\% & 32.8\% & 38\% & 27\% & 46.5\% & 48.5\% \\
		$\epsilon = 2/255$ & $>25$h & $344$s & $361$s & $32$s &  $58$s  \\  \hline
	\end{tabular}
	\caption{Images verified by different verifiers ($\%$, higher is better) on CIFAR CNN-B-Adv. Results, from \cite{cutting}, are not fully comparable with Tables \ref{table_beta},\ref{table_hybrid}.}
	\label{table_gcp}
	\vspace{-0.3cm}
\end{table}

Results on other networks from other verifiers (PRIMA \cite{prima}, SDP-FO \cite{SDPFI}, etc) were already reported \cite{crown} on other tested DNNs, with unfavorable comparison vs $\alpha,\beta$-CROWN. Further, we reported accuracy of NNenum \cite{nnenum}, Marabou \cite{Marabou,Marabou2}, respectively 4th, 3rd of the last VNNcomp'24 \cite{VNNcomp24}, as well as full MILP \cite{MILP} in Table \ref{table_complete}, showing that these verifiers are not competitive on complex (even small) instances. Concerning MnBAB \cite{ferrari2022complete}, we tested it in appendix, and it compares slightly unfavorably in time and accuracy towards $\alpha,\beta$-CROWN on CNN-B-Adv and {\em complex} MNIST DNNs at several time-out settings. Last, Pyrat \cite{pyrat} (2\textcolor{red}{nd} in the latest VNNComp) is not open source, which made running it impossible. Details can be found in the appendix. Also, ablation studies showing what each feature of Hybrid MILP is bringing can be found in the appendix.

>>>>

\begin{figure}[b!]
	\centering
%	\vspace*{-0.3cm}
	\includegraphics[height=7cm]{CNN-B-ADV_layer7_comparison_image85.png}.
%	\vspace*{-0.4cm}
	\caption{Average uncertainty of pMILP for neurons of the fourth layer of CNN-B-Adv, 
	when selecting between 0 and 30 ReLUs with different scoring functions.}
	\label{fig_table3}
\end{figure}

@@

Overall, {\sf SAS} is more accurate than GS, more accurate than Huang, more accurate than random choice of variables. {\sf SAS} significantly outperforms other solutions, 
with 2 times less binary variables for the same accuracy (10 vs 22) vs {\sf GS} and 6 times vs Huang \cite{DivideAndSlide} (4 vs 24).
Each node of the 4th layer displays a similar pattern, and the pattern is similar for different images.

@@

\begin{table}[t!]	
	\centering
	\caption{Comparison of different orderings with selection of ReLUs by {\sf SAS}.
	}
	\label{table.order}
	\begin{tabular}{|c||c c|c c|c c|c c|}
		\hline
		  Image 85& \multicolumn{2}{c|}{45 open ReLUs} & \multicolumn{2}{c|}{50 open ReLUs} 
		  & \multicolumn{2}{c|}{55 open ReLUs} & \multicolumn{2}{c|}{60 open ReLUs}\\ 
		 order & distance & time & distance & time & distance & time & distance & time \\
		\hline \hline
		Gurobi & 0.384 & 252s & 0.368&265s  &0.342 & 287s& 0.315  & 306s  \\ 
		{\sf SAS} (static) & 0.384 & 448s & 0.368&759s  &0.365 &971s & 0.355  & 971s \\
		{\sf GS} (static) & 0.384 & {\bf 251s} &0.368 &{\bf 253s}  & 0.342 &{\bf 270s} & 0.315 & {\bf 281s} \\ \hline
	\end{tabular}
%		\vspace{-0.6cm}
\end{table}

@@

\begin{figure}[b!]
%	\hspace*{-0.8cm}
	\includegraphics[scale=0.5]{CNN-B-ADV_layer7_comparison_image85_4methods.png}
	\caption{Distance to verify vs runtime: comparison between {\sf SAS}, GS, full MILP and Huang's method for different number of opened ReLUs / time-outs.}
	\label{fig555}
\end{figure}

{\sf SAS} (even {\sf GS}$=s_{FSB}$) is much more accurate than full MILP for every reasonable time-outs considered, with $>10$ times fastest runtime at better accuracy. The reason is that even the advanced MILP solver Gurobi struggles to sort out all ReLUs. It is thus particulary important to have accurate scoring functions as {\sf SAS}, in order to optimize both accuracy and runtime. Huang \cite{DivideAndSlide} is limited in ReLUs in the previous layer, reason why it is stuck at relatively poor accuracy. The accuracy/runtime curve from {\sf GS} is closer to {\sf SAS} than the number of nodes / accuracy curve of Fig. \ref{fig_table3}. This is because many ReLU nodes deemed important by {\sf GS} are not relevant for accuracy, but they are also not penalizing runtime. Still, {\sf SAS} is faster than {\sf GS} at  every accuracy. Importantly, {\sf SAS} is more deterministic in its runtime given a number of ReLU nodes, with half the variance in runtime over different output neurons compared with {\sf GS} for similar accuracy, which helps setting a number of ReLU nodes to open.

@@

\subsection{Ablation study: on usefulness of computing previous layers accurately}

We explore the usefulness of computing accurately bounds for each neuron inductively on the layers, 
even on small networks. For that, we consider MNIST $5 \times 100$, computing bounds for nodes of layer 3, comparing when bounds for neurons of layer 2
have been computed inaccurately using LP rather than with the more accurate (partial) MILP.

\begin{figure}[b!]
	\vspace*{-0.8cm}
	\includegraphics[scale=0.55]{Layer3_comparison_LP}.
	\caption{Comparison of accuracy in layer 3 when bounds for neurons in layer 2 is computed inaccurately using LP vs when bounds of layer 2 are computed accurately using pMILP.
	Time is using logscale.}
	\label{fig3LP}
\end{figure}

This experiment explains the rationale to use divide and conquer methodology, 
using many calls (one for each neuron) of pMILP with relatively small number $|X|$ of open ReLUs 
rather than few calls (one per output neuron) of pMILP with larger number $|X|$ of open nodes. 
The benefit is clear already from layer 3, obtaining much tighter bounds (lower uncertainty) when bounds for neurons in layer 2 have been computed accurately using pMILP.

@@

\begin{table}[b!]
	\centering
	\begin{tabular}{||l||c|c|c||c||c||}
		\hline \hline
		 & $\alpha,\beta$-CROWN & $\alpha,\beta$-CROWN & $\alpha,\beta$-CROWN & Refined & \bf Hybrid \\ 
		 Network & TO=10s & TO=30s & TO=2000s & $\beta$-CROWN & \bf MILP \\ 
		\hline
		MNIST 5$\times$100 & 57\% (6.9s) & 55\% (18.9s) & 50\% (1026s) & 13\% 
		(92s) & 13\% \bf (46s) \\ \hline
		MNIST 5$\times$200 & 50\% (6.5s) & 47\% (17s) & 46\% (930s) & 9\% (80s) & \bf 8\% (71s) \\ \hline
		MNIST 8$\times$100 & 63\% (7.2s) & 58\% (20s) & 58\% (1163s) & 21\% (102s) & \bf 15\% (61s) \\ \hline
		MNIST 8$\times$200 & 56\% (6.8s) & 55\% (18s) & 54\% (1083s) & 16\% (83s) & \bf 8\% (78s) \\ \hline
		MNIST 6$\times$500 & 53\% (6.4s) & 51\% (16s) & 50\% (1002s) & $-$ & 
		\bf 10\% (402s) \\ \hline
		CNN-B-Adv & 28\% (4.3s) & 22\% (8.7s) & 20\% (373s) & $-$ & \bf 11\% (417s) \\ \hline \hline
		ResNet & 0\% (2s) & 0\% (2s) & 0\% (2s) & $-$ & 0\% (2s) \\ \hline \hline
	\end{tabular}
	\caption{Undecided images ($\%$, {\em lower is better}) as computed by $\alpha,\beta$-CROWN, Refined $\beta$-CROWN and Hybrid MILP on 7 DNNs (average runtime per image). The 6 first DNNs are hard instances. The last DNN (ResNet) is an easy instance (trained using Wong to be easy to verify), provided for reference.
	}
	\label{table_hybrid}
%	\vspace*{-0.7cm}
	\end{table}

@@

\begin{table}[b!]
%	\vspace{-0.1cm}
	\centering
		\caption{Images verified by different verifiers ($\%$, higher is better) on CIFAR CNN-B-Adv. Results, from \cite{cutting}, are not fully comparable with Tables \ref{table_beta},\ref{table_hybrid}.}
	\label{table_gcp}
	\begin{tabular}{||c||c|c|c|c|c||}
		\hline
		Upper Bound & SDP-FO & PRIMA & refined $\beta$-CROWN & $\beta$-CROWN & GCP-CROWN \\  \hline
		65\% & 32.8\% & 38\% & 27\% & 46.5\% & 48.5\% \\
		$\epsilon = 2/255$ & $>25$h & $344$s & $361$s & $32$s &  $58$s  \\  \hline
	\end{tabular}
%	\vspace{-0.6cm}
\end{table}

Results on other networks from other verifiers (PRIMA \cite{prima}, SDP-FO \cite{SDPFI}, etc) were already reported \cite{crown} on other tested DNNs, with unfavorable comparison vs $\alpha,\beta$-CROWN. Further, we reported accuracy of NNenum \cite{nnenum}, Marabou \cite{Marabou,Marabou2}, respectively 4th, 3rd of the last VNNcomp'24 \cite{VNNcomp24}, as well as full MILP \cite{MILP} in Table \ref{table_complete}, showing that these verifiers are not competitive on complex (even small) instances. Concerning MnBAB \cite{ferrari2022complete}, and it compares slightly unfavorably in time and accuracy towards $\alpha,\beta$-CROWN on CNN-B-Adv and {\em complex} MNIST DNNs at several time-out settings. Last, Pyrat \cite{pyrat} (2nd in the latest VNNComp) is not open source, which made running it impossible. 
%Details can be found in the appendix. Also, ablation studies showing what each feature of Hybrid MILP is bringing can be found in the appendix.

### 2025-07-23 ATVA 2025/Final_version/relatedwork.tex
ATVA25

>>>>

Compared with \cite{atva}, which uses pMILP in an abstraction refinement loop, 
they iteratively call pMILP to obtain bounds for the same output neuron, opening more and more ReLUs. This scales only to limited size DNN (500 neurons), because of the fact that many ReLUs need to be open (and then Gurobi takes a lot of time) and the iterative nature which cannot be parallelized, unlike our method which scales up to 20.000 neurons.

%Instead, we call pMILP on each hidden neuron *once*, layer per layer (divide and conquer), with a fixed limited number of open ReLUs thanks to our novel SAS selection.
%pMILP opens the most important ReLUs for a given neuron.
